\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-20 paper.tex
Fix minor latex issues.

The scalability gains made by recent ARG inference methods have been, in part, due to inferring
lower levels of precision about recombination. Indeed, a common statement in the literature is
that modern tools such as \tsinfer \citep{kelleher2019inferring} and \relate \citep{speidel2019method}
construct not an ARG but "only" a sequence of local trees \[cite examples from \#38\]. While this statement
is correct inasmuch as neither method attempts to infer the explicit recombination events that cause
adjacent trees to differ, it is certainly not the case that they infer independent local trees.
In fact, both methods take advantage of genome-wide mutation information to construct genealogies,
and as a result nodes and edges persist across multiple local trees.
Their results can therefore be thought of as an inheritance graph, and represented in the gARG format.

To illustrate the varying levels of information that can be captured in a gARG,
Fig.~\ref{fig-inferred-args} presents graph representations of genealogies output by two tools that
\emph{do} explicitly infer recombination events, namely \kwarg \citep{ignatieva2021kwarg} and \argweaver
\citep{rasmussen2014genome}, and two which do not (tsinfer and Relate). All inferences are based on the
same underlying dataset from \citet{kreitman1983nucleotide}, a standard benchmark in the classical
ARG literature, for which the minimum number of recombination breakpoints in the absence of recurrent
mutation is 7~\citep{song2003parsimonious}. This dataset comprises 43 biallelic SNPs spanning 2.4Kb of the
\emph{D.~melanogaster} ADH locus on chromosome 2L. Where required for inference purposes we
assume a mutation rate of \SI{5.49e-9}{site^{-1}generation^{-1}} and a recombination rate of
\SI{8.4e-9}{site^{-1}generation^{-1}} [CITE stdpopsim] and a constant effective population size of 1000,000.
The version numbers of inference software, together with the specific inference parameters
needed to reproduce the ARGs, are available from our GitHub repository.

Note that both \kwarg and \argweaver produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args} A\& B are arbitrarily selected examples).
This stems from the fact that \kwarg searches the space of plausible ARGs,
outputting minimal ones using heuristics, and ARGweaver samples from the posterior distribution over
ARGs given the data and a fixed model (the SMC$\prime$, an approximation of the full coalescent
with recombination). In contrast, \tsinfer and \relate (Fig.~\ref{fig-inferred-args} C \& D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is a important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver (Fig.~\ref{fig-inferred-args} B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).

A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args} A \& B, each recombination node has exactly 2 parents,
corresponding to a single recombination event with a specific breakpoint. Assuming that each
breakpoint occurs a a different location (the "infinite breakpoints" model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args} C \& D have ARG nodes with greater
than two parents, indicating deliberate imprecision: these are cases where multiple recombination events
have been ``pushed down'' onto a single more recent node. Note, however, that the number
of breakpoints in these two ARGs is smaller than the number of ARG nodes with multiple parents,
because those nodes frequently share breakpoint positions. The sharing of breakpoint locations
between multiple-parent nodes indicates that in \tsinfer and \relate ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves.

Shading by coalescent span reveals that many non-recombinant (green) nodes are
``locally unary'' (see section \ref{sec-locally-unary-edges}). In other words,
in some regions of the genome, these nodes are present in the ancestry but do
not correspond to coalescence. As argued previously, this can provide potentially useful
information about shared ancestry. Note however, that such unary segments are not inferred
in the \relate ARG: other than the samples it has entirely solid nodes.
Only the \tsinfer results have nodes with more than two children (polytomies);
in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
uncertainty in the order of coalescent events, in an equivalent
manner to the lack of precision in recombination nodes with more
than two parents. The clearest example is the polytomy involving the nodes
\noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}, which happen to
have identical sequences. \kwarg, \argweaver, and \relate require
bifurcating trees, so are forced
to pick an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args} A and Fig.~\ref{fig-inferred-args} B disagree on the
relationship between these samples). TsThe number of nodes in Fig.~\ref{fig-inferred-args} C
that have more than two children also hints at considerable uncertainty in
the order of branching elsewhere in the ARG.
% TODO add explanation of the bottom row.
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
%The example of successive simplification of a simulated ARG
%in Fig.~\ref{fig-simplification} illustrates the point
%that is is possible to \emph{store} arbitrary levels of precision
%about recombination in a gARG.
%It is also possible to \emph{infer} various levels of precision about
%recombination.
%To illustrate this, and to demonstrate the structural
%heterogeneity of the ARGs inferred  by current methods,
%Fig.~\ref{fig-inferred-args} shows the results of ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
%for the classical \citet{kreitman1983nucleotide} benchmark dataset
%using  \kwarg~\citep{ignatieva2021kwarg},
%\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
%\tsinfer~\citep{kelleher2019inferring},
%and \relate~\citep{speidel2019method}.
% While there is some consistency among the inferred ARGs (such as the samples
% FR-F, WA-F and Af-F being grouped into the same clade), there
% are also considerable differences.
%The minimum number
%of recombination breakpoints required in the absence of recurrent mutation
%is 7 for this dataset~\citep{song2003parsimonious},
%and the number of inferred breakpoints
%under the input mutation and recombination rates
%varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
%The ARGs inferred by the different methods also have major differences
%in how recombination information is represented.
%Essentially,
%\kwarg\ and \argweaver\ infer eARGs that contain
%specific recombination \emph{events} for each breakpoint.
%On the other hand, \tsinfer\ and
%\relate\ do not infer events but rather
%infer gARGs with by differing levels of precision about recombination.

>>>>

The scalability gains made by recent ARG inference methods have been, in part, due to inferring
lower levels of precision about recombination. Indeed, a common statement in the literature is
that modern tools such as \tsinfer~\citep{kelleher2019inferring} and \relate~\citep{speidel2019method}
construct not an ARG but ``only'' a sequence of local 
trees [TODO cite examples from \#38]. While this statement
is correct inasmuch as neither method attempts to infer the explicit recombination events that cause
adjacent trees to differ, it is certainly not the case that they infer independent local trees.
In fact, both methods take advantage of genome-wide mutation information to construct genealogies,
and as a result nodes and edges persist across multiple local trees.
Their results can therefore be thought of as an inheritance graph, and represented in the gARG format.

To illustrate the varying levels of information that can be captured in a gARG,
Fig.~\ref{fig-inferred-args} presents graph representations of genealogies output by two tools that
\emph{do} explicitly infer recombination events, namely \kwarg~\citep{ignatieva2021kwarg} and 
\argweaver~\citep{rasmussen2014genome}, 
and two which do not (\tsinfer\ and \relate). 
All inferences are based on the
same underlying dataset from \citet{kreitman1983nucleotide}, a standard benchmark in the classical
ARG literature, for which the minimum number of recombination breakpoints in the absence of recurrent
mutation is 7~\citep{song2003parsimonious}. This dataset comprises 43 biallelic SNPs spanning 2.4Kb of the
\emph{D.~melanogaster} ADH locus on chromosome 2L. Where required for inference purposes we
assume a mutation rate of 
% per site per generation may be easier to follow?
\SI{5.49e-9}{site^{-1}generation^{-1}} and a recombination rate of
\SI{8.4e-9}{site^{-1}generation^{-1}} [CITE stdpopsim] and a constant effective population size of 
1,000,000.
The version numbers of inference software, together with the specific inference parameters
needed to reproduce the ARGs, are available from our GitHub repository.

Note that both \kwarg\ and \argweaver\ produce many ARGs
(those shown in Fig.~\ref{fig-inferred-args} A\& B are arbitrarily selected examples).
This stems from the fact that \kwarg\ searches the space of plausible ARGs,
outputting minimal ones using heuristics, and \argweaver\ samples 
from the posterior distribution over
ARGs given the data and a fixed model (the SMC$\prime$, an approximation of the full coalescent
with recombination) [TODO CITE]. 
In contrast, \tsinfer\ and \relate\ (Fig.~\ref{fig-inferred-args} C \& D) are
deterministic methods, aiming to find a single ``best guess'' ARG.
The ability of a gARG to encode the effects of recombination, rather
than having to precisely specify each event, is a important contributor to this aim,
as most recombination events cannot be uniquely identified from the data.
Indeed, there is considerable variation in the number of recombinations inferred by
different methods, with e.g.~\argweaver\ (Fig.~\ref{fig-inferred-args} B) suggesting
far more than are required under minimal parsimony assumptions (Fig.~\ref{fig-inferred-args} A).

A rough feel for the amount of recombination in each ARG is given by the distribution
of number of parents per node, as highlighted by the node colour scheme.
In Fig.~\ref{fig-inferred-args} A \& B, each recombination node has exactly 2 parents,
corresponding to a single recombination event with a specific breakpoint. Assuming that each
breakpoint occurs a a different location (the "infinite breakpoints" model), then the
number of breakpoints should exactly equal the number of 2-parent (brown) nodes in the plot.
In contrast, Figure Fig.~\ref{fig-inferred-args} C \& D have ARG nodes with greater
than two parents, indicating deliberate imprecision: these are cases where multiple recombination events
have been ``pushed down'' onto a single more recent node. Note, however, that the number
of breakpoints in these two ARGs is smaller than the number of ARG nodes with multiple parents,
because those nodes frequently share breakpoint positions. The sharing of breakpoint locations
between multiple-parent nodes indicates that in \tsinfer\ and \relate\ ARGs it may be
difficult to condense each transition between trees into a set of simple SPR moves.

Shading by coalescent span reveals that many non-recombinant (green) nodes are
``locally unary'' (see section \ref{sec-locally-unary-edges}). In other words,
in some regions of the genome, these nodes are present in the ancestry but do
not correspond to coalescence. As argued previously, this can provide potentially useful
information about shared ancestry. Note however, that such unary segments are not inferred
in the \relate\ ARG: other than the samples it has entirely solid nodes.
Only the \tsinfer\ results have nodes with more than two children (polytomies);
in Appendix \ref{sec-cell-lineages-and-args} we argue that this reflects
uncertainty in the order of coalescent events, in an equivalent
manner to the lack of precision in recombination nodes with more
than two parents. The clearest example is the polytomy involving the nodes
\noderef{Fr-F}, \noderef{Wa-F}, and \noderef{Af-F}, which happen to
have identical sequences. \kwarg, \argweaver, and \relate\ require
bifurcating trees, so are forced
to pick an arbitrary order of branching
(hence Fig.~\ref{fig-inferred-args} A and Fig.~\ref{fig-inferred-args} B disagree on the
relationship between these samples). 
The number of nodes in Fig.~\ref{fig-inferred-args} C
that have more than two children also hints at considerable uncertainty in
the order of branching elsewhere in the ARG.
% TODO add explanation of the bottom row.
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/386
%The example of successive simplification of a simulated ARG
%in Fig.~\ref{fig-simplification} illustrates the point
%that is is possible to \emph{store} arbitrary levels of precision
%about recombination in a gARG.
%It is also possible to \emph{infer} various levels of precision about
%recombination.
%To illustrate this, and to demonstrate the structural
%heterogeneity of the ARGs inferred  by current methods,
%Fig.~\ref{fig-inferred-args} shows the results of ARG inference
% To illustrate these points and to emphasise the heterogeneity
% of the ARGs inferred by current methods, we inferred ARGs for the
%for the classical \citet{kreitman1983nucleotide} benchmark dataset
%using  \kwarg~\citep{ignatieva2021kwarg},
%\argweaver~\citep{rasmussen2014genome,hubisz2020inference},
%\tsinfer~\citep{kelleher2019inferring},
%and \relate~\citep{speidel2019method}.
% While there is some consistency among the inferred ARGs (such as the samples
% FR-F, WA-F and Af-F being grouped into the same clade), there
% are also considerable differences.
%The minimum number
%of recombination breakpoints required in the absence of recurrent mutation
%is 7 for this dataset~\citep{song2003parsimonious},
%and the number of inferred breakpoints
%under the input mutation and recombination rates
%varies from 2 (\relate) to 17 (\argweaver).
% TODO this needs a bit of rewriting with the rest of the paper in mind
% kwarg and argweaver are inferring eARGs, but we need to be careful
% not to confuse people.
%The ARGs inferred by the different methods also have major differences
%in how recombination information is represented.
%Essentially,
%\kwarg\ and \argweaver\ infer eARGs that contain
%specific recombination \emph{events} for each breakpoint.
%On the other hand, \tsinfer\ and
%\relate\ do not infer events but rather
%infer gARGs with by differing levels of precision about recombination.
